Understanding the role of different proxy selection methods in High-Dimensional Propensity Score Analysis

### 2024-09-01 Writeup/manuscript.tex
appFigures

\usepackage{hyperref}
\usepackage[utf8]{inputenc}
\usepackage{capt-of}
\usepackage{booktabs}
\usepackage{amssymb}
\usepackage{threeparttable}
\usepackage{float}
%\floatplacement{figure}{H}
%\floatplacement{table}{H}
\usepackage{lipsum,caption}
\geometry{
  a4paper,
  left=1in,
  right=1in,
  top=1in,
  bottom=1in,
  includeheadfoot
}
\usepackage{setspace}
%\usepackage{lineno}
%\linenumbers
\usepackage{siunitx}
\sisetup{
  mode = match,
  propagate-math-font = true,
  reset-math-version = false,
  reset-text-family = false,
  reset-text-series = false,
  reset-text-shape = false,
  text-family-to-math = true,
  text-series-to-math = true
}
\doublespacing
% Disable explicit page breaks in LaTeX
\let\clearpage\relax

@@

\title[Deep Learning in Propensity Score Analysis]{Challenges of
Univariate Proxy Variable Screening in High-Dimensional Propensity Score
Analysis - An Argument for Multivariate Approaches}

@@

\author*[3]{\fnm{Yang} \sur{Lei} }

@@

\abstract{\textbf{Purpose}:

\textbf{Methods:}

\textbf{Results:}

\textbf{Conclusion:}}

@@

\textbf{Aim}:

@@

\textbf{Right Heart Catheterization dataset}:

\textbf{Plasmode simulation}: See Table \ref{table:scenarios} for the
description of the scenarios under consideration.

\begin{table}[ht]
\centering
\caption{Simulation Scenarios for plasmode simulation based on the Right Heart Catheterization (RHC) study.}
\label{table:scenarios}
\begin{tabular}{lcccc}
  \toprule
  \textbf{Plasmode Simulation Scenario} & \textbf{Exposure} & \textbf{Outcome} & \textbf{True} & \textbf{Sample}\\
  \textbf{} & \textbf{Prevalence} & \textbf{Prevalence} & \textbf{Odds Ratio} & \textbf{Size}\\
  \midrule
  (i) Frequent Exposure and Outcome (Base) & 30\% & 30\% & 1 & 3,500 \\
  (ii) Rare Exposure and Frequent Outcome & 5\% & 30\% & 1 & 3,500 \\
  (iii) Frequent Exposure and Rare Outcome & 30\% & 5\% & 1 & 3,500 \\
  \bottomrule
\end{tabular}
\end{table}

\textbf{True data generating mechanism used in plasmode simulation}:

\textbf{Performance measures}: From this simulation, we derived several
performance metrics: (1) bias, (2) average model standard error (SE; the
average of estimated SEs obtained from a model over repeated samples),
(3) empirical SE (the standard deviation of estimated treatment effects
across repeated samples), (4) MSE, (5) coverage probability of 95\%
confidence intervals, (6) bias-eliminated coverage, and (7) Zip plot
\citep{morris2019using, white2023check}.

@@

@@

@@

@@

@@

\renewcommand\refname{References}
\bibliography{mybibfile.bib}

>>>>

\usepackage{natbib}
\usepackage{hyperref}
\usepackage[utf8]{inputenc}
\usepackage{capt-of}
\usepackage{booktabs}
\usepackage{amssymb}
\usepackage{threeparttable}
\usepackage{float}
%\floatplacement{figure}{H}
%\floatplacement{table}{H}
\usepackage{lipsum,caption}
\geometry{
  a4paper,
  left=1in,
  right=1in,
  top=1in,
  bottom=1in,
  includeheadfoot
}
\usepackage{setspace}
%\usepackage{lineno}
%\linenumbers
\usepackage{siunitx}
\sisetup{
  mode = match,
  propagate-math-font = true,
  reset-math-version = false,
  reset-text-family = false,
  reset-text-series = false,
  reset-text-shape = false,
  text-family-to-math = true,
  text-series-to-math = true
}
\doublespacing
% Disable explicit page breaks in LaTeX
\let\clearpage\relax

@@

\title[High-Dimensional Propensity Score]{Understanding the role of
different proxy selection methods in High-Dimensional Propensity Score
Analysis}

@@

\author[3]{\fnm{Yang} \sur{Lei} }

@@

\abstract{\textbf{Purpose}: \textbf{Methods:} \textbf{Results:}
\textbf{Conclusion:}}

@@

\textbf{Aim}: The aim of this research is to systematically evaluate and
compare different proxy selection methods within the context of
high-dimensional propensity score (hdPS) analysis. Specifically, the
study focuses on assessing how these methods, including alternative
variable selection approaches, perform in selecting proxy variables for
confounding adjustment compared to the traditional Bross method within
the hdPS framework. We seek to determine whether these alternative
methods offer superior performance in estimating treatment effects
compared to the default Bross formula.

@@

\textbf{Motivating Example}: To explore the relationship between obesity
and the risk of diabetes, we revisited this association using data from
three cycles of the National Health and Nutrition Examination Survey
(NHANES) covering the years 2013-2014, 2015-2016, and 2017-2018
\citep{karim2024high}. This analysis was informed by a thorough review
of the existing literature
\citep{saydah2014trends, liu2013association, kabadi2012joint, ostchega2012abdominal}.
To identify relevant covariates, we constructed a causal diagram based
on established causal inference principles \citep{greenland1999causal}.
The covariates included in our analysis were carefully selected and
categorized into Demographic, Behavioral, Health History,
Access-related, and Laboratory variables \citep{karim2024high}. While
most of these variables were binary or categorical, the Laboratory
variables were continuous.

\textbf{Plasmode simulation}: To rigorously assess the performance of
the methods under consideration, we employed a plasmode simulation
framework, which is particularly well-suited for reflecting real-world
data structures and complexities \citep{franklin2014plasmode}. This
approach was modeled after the analytic dataset derived from NHANES and
involved resampling from the observed covariates and exposure
information (i.e., obesity) without altering them. By mirroring key
aspects of an actual epidemiological study, this simulation framework
offers a significant advantage over traditional Monte Carlo simulations,
which often rely on idealized assumptions.

\textbf{Simulation scenarios under consideration}: Our plasmode
simulation was conducted over 500 iterations. For the base simulation
scenario, we set the prevalence of exposure (obesity) and the event rate
(diabetes) at 30\%, with a true odds ratio (OR) parameter of 1,
corresponding to a risk difference (RD) of 0. Each simulated dataset had
a sample size of 3,000 participants. The description of other scenarios
under consideration is provided in Table \ref{table:scenarios}.

\begin{table}[ht]
\centering
\caption{Overview of Plasmode Simulation Scenarios Reflecting Varying Exposure and Outcome Prevalences Based on National Health and Nutrition Examination Survey (NHANES) Data Cycles (2013-2018)}
\label{table:scenarios}
\begin{tabular}{lcccc}
  \toprule
  \textbf{Plasmode Simulation Scenario} & \textbf{Exposure} & \textbf{Outcome} & \textbf{True} & \textbf{Sample}\\
  \textbf{} & \textbf{Prevalence} & \textbf{Prevalence} & \textbf{Odds Ratio} & \textbf{Size}\\
  \midrule
  (i) Frequent Exposure and Outcome (Base) & 30\% & 30\% & 1 & 3,000 \\
  (ii) Rare Exposure and Frequent Outcome & 5\% & 30\% & 1 & 3,000 \\
  (iii) Frequent Exposure and Rare Outcome & 30\% & 5\% & 1 & 3,000 \\
  \bottomrule
\end{tabular}
\end{table}

\textbf{True Data Generating Mechanism Used in Plasmode Simulation}: The
primary goal of this study is to evaluate various variable selection
methods under realistic conditions. To achieve this, we formulated the
outcome data based on a specific model specification that incorporates
both exposure and covariates, including investigator-specified and proxy
variables. The model specification consists of three key components:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  \emph{Investigator-Specified Covariates}: We retained the original
  investigator-specified covariates, which were either binary or
  categorical, reflecting how real-world studies typically operate.
\item
  \emph{Transformation of Laboratory Variables}: In real-world studies,
  it is common for analysts to lack precise knowledge of the true model
  specification. To simulate this uncertainty, we transformed the
  continuous laboratory variables using complex functions such as
  logarithmic, exponential, square root, polynomial transformations, and
  interactions. This reflects the challenges analysts face in correctly
  specifying models when dealing with continuous data.
\item
  \emph{Inclusion of Proxy Variables}: Real-world studies often deal
  with unmeasured confounding, which researchers attempt to mitigate by
  adding proxy variables. However, when a large number of proxies are
  added, some may act as noise variables, contributing little to the
  analysis. To simulate this, we selected only those binary proxy
  covariates (referred to as recurrence covariates in hdPS terminology)
  that had a relative risk (RR) of less than 0.8 or greater than 1.2
  concerning the outcome. Out of 143 proxy covariates, 94 met this
  criterion and were included in calculating a simple comorbidity burden
  measure. The remaining 49 covariates were excluded from this
  calculation and considered noise. This comorbidity burden measure was
  then incorporated into our model specification for generating the
  plasmode data.
\end{enumerate}

\textbf{Performance Measures}: From this simulation, we derived several
performance metrics to evaluate the effectiveness of the methods under
consideration: (1) bias, (2) average model standard error (SE; the
average of estimated SEs obtained from a model over repeated samples),
(3) empirical SE (the standard deviation of estimated treatment effects
across repeated samples), (4) mean squared error (MSE), (5) coverage
probability of 95\% confidence intervals, (6) bias-corrected coverage,
and (7) Zip plot \citep{morris2019using, white2023check}.

@@

The comparison between the data generation process and the analysis
process reveals two key differences: (i) The data generation used
transformed laboratory variables, whereas the analysis was conducted
using only the original laboratory variables. (ii) The data generation
employed a simple sum of selected proxy variables (sum of 94 proxy
covariates), while the analysis included all proxy variables (143 binary
proxies), with 49 of these acting as noise variables. These differences
help us assess how the proxy variable selection methods handle model
misspecification and the presence of noise variables.

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  \textbf{Kitchen sink model}: This is a base model for comparison,
  where no variable selection approaches were used. All
  investigator-selected features and all proxy variables were used to
  model \citep{karim2018can}.
\item
  \textbf{Bross formula}: The Bross formula is a statistical method used
  to calculate the bias introduced by not adjusting for a covariate
  \citep{bross1966spurious}. In hdPS analysis, this formula was
  originally applied to each proxy variable to measure and rank the
  potential bias if the covariate were not adjusted for. In our
  analysis, the 100 proxies with the highest bias rankings are selected
  for further modeling \citep{schneeweiss2009high, wyss2018erratum}.
\item
  \textbf{Least Absolute Shrinkage and Selection Operator (LASSO)}:
  LASSO is a variable selection technique that limits the number of
  variables by adding a penalty term to the regression model.
  Cross-validation (CV) is used in LASSO to identify variables with
  non-zero coefficients in the best model by optimizing the penalty
  value
  \citep{franklin2015regularized, schneeweiss2017variable, karim2018can}.
\item
  \textbf{Hybrid of hdPS and LASSO}: Instead of relying solely on LASSO
  for variable selection, a hybrid approach combines the Bross formula
  and LASSO. First, hdPS variables are selected using the hdPS algorithm
  (e.g., the top 100), and then LASSO is applied to further refine the
  selection \citep{karim2018can, franklin2015regularized}.
\item
  \textbf{Elasticnet}: Elastic Net is an extension of LASSO that
  includes an additional penalty term to handle multicollinearity by
  grouping correlated features and selecting the most representative
  ones \citep{karim2018can}.
\item
  \textbf{Random Forest}: The Random Forest (RF) algorithm is an
  ensemble learning method that constructs multiple decision trees to
  perform classification \citep{breiman2001random}. It calculates the
  importance of each proxy variable based on the decrease in impurity or
  Gini importance, providing a ranking of the proxies. The top 100
  variables from this ranking are manually selected for further modeling
  \citep{schneeweiss2017variable}.
\item
  \textbf{XGBoost}: XGBoost is a gradient boosting algorithm used to
  optimize machine learning models \citep{chen2016xgboost}. It builds
  decision trees that make splits based on maximum impurity reduction,
  and it assigns an importance score to each proxy variable by
  calculating the mean decrease in impurity
  \citep{xiao2024interpretable}.
\item
  \textbf{Stepwise}: Stepwise selection is a progressive feature
  selection method that can proceed in two directions---forward or
  backward---based on the maximum adjusted R-squared. We have
  implemented two versions: (a) Forward selection (FS) starts with an
  initial model (e.g., including all investigator-selected features) and
  adds proxies to the model one at a time. (b) Backward elimination (BE)
  starts with a full model (e.g., all investigator-selected features and
  all proxy variables) and removes features one at a time based on their
  contribution to the model.
\item
  \textbf{Genetic algorithm (GA)}: GA is an evolutionary algorithm
  inspired by the theory of natural selection
  \citep{holland1975adaptation}. It operates by evolving offspring from
  a population of the fittest individuals over several generations,
  evaluating and selecting the best combination of features or variables
  that maximize prediction accuracy.
\end{enumerate}

@@

\begin{figure}[th]

{\centering \includegraphics[width=1\linewidth,]{figures/metric_comparison_Bias}

}

\caption{Comparison of Bias Across Different Methods in hdPS Analysis}\label{fig:unnamed-chunk-1}
\end{figure}

\begin{figure}[th]

{\centering \includegraphics[width=1\linewidth,]{figures/metric_comparison_Coverage}

}

\caption{Comparison of Coverage Probability Across Different Methods in hdPS Analysis}\label{fig:unnamed-chunk-2}
\end{figure}

@@

\emph{Here we include full data analysis (with some summary results like
exposure and outcome prevalence, and sample size) and report OR and RD.
Also mention how many proxies were chosen (add in the picture of RD and
OR; side by side for each method, ordered my magnitude of RD), and how
many were in common with hdPS (add table).}

\begin{figure}[th]

{\centering \includegraphics[width=1\linewidth,]{manuscript_files/figure-latex/unnamed-chunk-3-1}

}

\caption{Figure presenting a comparison of Risk Differences (RD) and Odds Ratios (OR) with 95\% confidence intervals for different methods used to evaluate the association between obesity and diabetes risk. The analysis is based on data from the National Health and Nutrition Examination Survey (NHANES) for the years 2013-2018. Methods are arranged by the number of variables used in the models.}\label{fig:unnamed-chunk-3}
\end{figure}

Table \ref{tab:method-comparison} presents a pairwise comparison of the
number of proxy features shared between different variable selection
methods used in the analysis. Each cell in the table indicates the count
of common proxy variables selected by the method in the corresponding
row and column. The diagonal cells, where the row and column methods are
the same, represent the total number of proxy variables selected
exclusively by each method.

\begin{table}[htbp]
\centering
\caption{Comparison of variable overlap of selected proxies across different methods used to evaluate the association between obesity and diabetes}
\label{tab:method-comparison}
\begin{tabular}{lccccccccc}
\toprule
 & \textbf{Bross} & \textbf{Hybrid} & \textbf{LASSO} & \textbf{Elasticnet} & \textbf{GA} & \textbf{XGBoost} & \textbf{RF} & \textbf{FS} & \textbf{BE} \\
\midrule
\textbf{Bross formula} & 100 & & & & & & & & \\
\textbf{Hybrid (Bross and LASSO)} & 49 & 49 & & & & & & & \\
\textbf{LASSO} & 47 & 47 & 60 & & & & & & \\
\textbf{Elasticnet} & 54 & 48 & 60 & 69 & & & & & \\
\textbf{Genetic algorithm (GA)} & 44 & 28 & 36 & 40 & 64 & & & & \\
\textbf{XGBoost} & 38 & 24 & 28 & 30 & 25 & 48 & & & \\
\textbf{Random Forest (RF)} & 72 & 37 & 42 & 50 & 36 & 48 & 100 & & \\
\textbf{Forward selection (FS)} & 45 & 41 & 51 & 54 & 35 & 25 & 43 & 59 & \\
\textbf{Backward elimination (BE)} & 45 & 41 & 51 & 54 & 35 & 25 & 43 & 59 & 59 \\
\bottomrule
\end{tabular}
\end{table}

@@

\emph{Report computing time for the Real-world analysis for each method.
(add ordered table)}

@@

\renewcommand\refname{References}
\bibliography{mergedbibliography.bib}
